Cloudbreak: A MapReduce Algorithm for Detecting Genomic Structural Variation from Complete Paired-End Sequence Data Sets

### 2013-01-03 manuscript/recomb.tex
update to manuscript

Genomic structural variations such as deletions, insertions, and inversions of DNA, are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.

@@

RP methods have begun to consider larger and larger data sets. Although the first RP methods used only unambiguously, discordantly mapped read pairs in their analyses, it has been demonstrated that including multiple mappings of discordant read pairs in their analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}. Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.

@@

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We rank the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions.

We fit the means and of the GMM using the Expectation-Maximization algorithm. Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and the library to have mean fragment size $\mu$ with standard deviation $\sigma$. We initialize the two components to have means $\mu$ and $\bar{Y}$ and standard deviation $\sigma$, and set $\alpha = .5$. In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$. We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions. In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$. We repeat E and M steps until convergence or a maximum number of steps has been taken.

Our features for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.

\item[\sc{PostProcess}] We convert our features along the genome to deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold. To eliminate noise we apply a median filter with window size 5. We end regions when $\mu'$ changes by more than 60bp, discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
\end{description}

>>>>

Genomic structural variations such as deletions, insertions, and inversions of DNA are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \autocite{Mills:2011p1611, Conrad:2010ja}. However, detection of structural variations with current high-throughput sequencing technology remains a difficult problem, with limited concordance between available algorithms and high false discovery rates \autocite{Mills:2011p1611}.

@@

RP methods have begun to consider larger and larger data sets. Although the first RP methods used only unambiguously discordantly mapped read pairs in their analyses, it has been demonstrated that including multiple mappings of discordant read pairs in the analysis improves sensitivity in repetitive and hard-to-map regions of the genome \autocite{Hormozdiari:2009p284,Quinlan:2010gf}. More recently, several RP approaches have considered concordant read pairs, either to integrate RD signals for improved sensitivity and specificity \autocite{Sindi:2012kk,Michaelson:2012fj,Chiara:2012ey}, or to eliminate the thresholds necessary to separate concordant from discordant mappings and be able to detect smaller events \autocite{Marschall:2012ek}. Even so, however, most methods use only a limited number of ambiguous discordant mappings per read pair, because of the storage and computational requirements necessary to process all or most ambiguous mappings of each read pair in a high-coverage data set.

@@

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We rank the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions. To estimate the probability that the mapping is correct, we approximate the probability $p^{k}_c$ of the alignment of each end being correct by:

\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]

And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.

We fit the means and variances of the GMM using the Expectation-Maximization algorithm. Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$. We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both to $\sigma$, and set $\alpha = .5$. In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$. We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions. In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\sum_m{\gamma_{m,2} + y_m}$. We treat the variance as fixed and do not update it, since under our assumptons the standard deviation of each component should always be $\sigma$. We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.

The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.

\item[\sc{PostProcess}] We convert our features along the genome to deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold. To eliminate noise we apply a median filter with window size 5. We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
\end{description}
